R and Python clients for GBIF species occurrence data

### 2016-08-05 manuscript.Rmd
tried to fix front page for peerj preprint specs still not quite right fixed some wkt stuff improved use cases to not take so much time

---
layout: review, 11pt
linenumbers: true
title: "R and Python clients for GBIF species occurrence data"
author:
  - name: Scott Chamberlain
    affiliation: cstar
    email: scott(at)ropensci.org
    footnote: Corresponding author
  - name: Carl Boettiger
    affiliation: boettig
    email: carl(at)ropensci.org
address:
  - code: cstar
    address: |
      University of California, Berkeley, CA, USA
  - code: boettig
    address: |
      University of California, Berkeley, CA, USA
abstract: |
      1. xxx

      2. xxx

      3. xxx

      4. xxxx

@@

```{r compile-settings, include=FALSE}
library("methods")
library("knitr")
opts_chunk$set(
  tidy = FALSE,
  warning = FALSE,
  message = FALSE,
  cache = 1,
  comment = '#>',
  collapse = TRUE,
  verbose = TRUE
)
basename <- gsub(".Rmd", "", knitr:::knit_concord$get('infile'))
opts_chunk$set(fig.path = paste("components/figure/", basename, "-", sep=""),
               cache.path = paste("components/cache/", basename, "/", sep=""))
```

@@

All this organization matters because you can navigate through GBIF data through all
these organizational levels.

@@

## The rgbif package

@@

## The pygbif library

`pygbif` [@pygbif] is a Python library for working with GBIF data in
the Python programming environment [@python]. Python is a general
purpose programming language used widely in all sectors, and for
all parts of software development including server and client
side use cases. Python is used exclusively in some scientific
disciplines (e.g., astronomy), and has partial usage in other
disciplines. A Python client for GBIF data is an important tool
given the even wider usage of Python than R, though maybe slightly
less than R for ecology/biology.

@@

The `pygbif` library is less mature and complete than the R package. It also
uses an [MIT license][mit] to maximize use everywhere. `pygbif` is developed
publicly on GitHub at [https://github.com/sckott/pygbif][ghpygbif], where
development versions of the package can be installed, and bugs and feature
requests reported. Stable versions of `pygbif` can be installed from
[pypi][pypipygbif], the distribution network for Python libraries.

@@

The output of these examples, except one, aren't shown, but all run correctly.

@@

```{r eval=FALSE}
# well known text
occ_search(geometry = 'POLYGON((30.1 10.1, 10 20, 20 40, 40 40, 30.1 10.1))', limit = 20)
# bounding box
occ_search(geometry = c(-125.0,38.4,-121.8,40.9), limit = 20)
```

@@

@@

@@

The `rgbif` and `pygbif` libraries provide programmatic interfaces to GBIF's
application programming interface (API) - a powerful tool for making
research using species occurrence data reproducible. In fact,
the `rgbif` package has already been used in 22 scholarly publications
(as of 2015-11-14).

@@

The `pygbif` library is still in early development, and will
greatly benefit from any feedback and use cases.

One area of focus in the future is to attempt to solve many use
cases that have been brought up with respect to GBIF data. For
example, some specimens are included in GBIF that are located
in botanical gardens. For many research questions, researchers are
interested in "wild" type occurrences, not those in human curated
scenarios. Making removal of these occurrences easy would be
very useful, but is actually quite a hard problem.

@@

All scripts and data used in this paper can be found in the permanent
data archive Zenodo under the digital object identifier (DOI). This DOI
corresponds to a snapshot of the GitHub repository at
[github.com/sckott/msrgbif](https://github.com/sckott/msrgbif). Software
can be found at [https://github.com/ropensci/rgbif](https://github.com/ropensci/rgbif),
under an MIT license.

>>>>

---
layout: review, 11pt
linenumbers: true
title: "R and Python clients for GBIF species occurrence data"
author:
  - name: Scott Chamberlain
    affiliation: cstar
    email: scott(at)ropensci.org
    footnote: Corresponding author
  - name: Carl Boettiger
    affiliation: boettig
    email: carl(at)ropensci.org
address:
  - code: cstar
    address: |
      rOpenSci, Museum of Paleontology, University of California, Berkeley, CA, USA
  - code: boettig
    address: |
      rOpenSci, Department of Enivornmental Science, Policy and Management, University of California, Berkeley, CA, USA
abstract: |
      Corresponding Author:

      Scott Chamberlain

      rOpenSci, Museum of Paleontology, University of California, Berkeley, CA, USA

      Email address: [scott@ropensci.org](mailto:scott@ropensci.org)

      \newpage

      Background. The number of individuals of each species in
        a given location forms the basis for many sub-fields of ecology and
        evolution. Data on individuals, including which species, and
        where they're found, can be used for a large number of research
        questions. Global Biodiversity Information Facility (hereafter, GBIF)
        is the largest of these. Programmatic clients for GBIF would
        make research dealing with GBIF data much easier.

      Methods. We have developed a client to access GBIF data for each
        of the R and Python programming languages: `rgbif` and `pygbif`.

      Results. For both clients, we describe their design and utility,
        and demonstrate many use cases.

      Discussion. Programmatic acccess to GBIF will facilitate more open
        and reproducible science. We hope that all scientists using GBIF
        data will do so with one of these two clients.

@@

```{r compile-settings, include=FALSE}
library("methods")
library("knitr")
opts_chunk$set(
  tidy = FALSE,
  warning = FALSE,
  message = FALSE,
  cache = 1,
  comment = '#>',
  collapse = TRUE,
  verbose = TRUE
)

basename <- gsub(".Rmd", "", knitr:::knit_concord$get('infile'))
opts_chunk$set(fig.path = paste("components/figure/", basename, "-", sep=""),
               cache.path = paste("components/cache/", basename, "/", sep=""))

# tibble options
options(tibble.max_extra_cols = 10)
```

\newpage

@@

This organization matters because you can navigate through GBIF data through
these various organizational levels.

@@

## rgbif

@@

## pygbif

`pygbif` [@pygbif] is a Python library for working with GBIF data in
the Python programming environment. Python is a general
purpose programming language used widely in all sectors, and for
all parts of software development including server and client
side use cases. Python is used exclusively in some scientific
disciplines (e.g., astronomy), and has partial usage in other
disciplines. A Python client for GBIF data is an important tool
given the even wider usage of Python than R, though maybe slightly
less than R for ecology/biology disciplines.

@@

The `pygbif` library is less mature and complete than the R package. It also
uses an [MIT license][mit] to maximize use everywhere. `pygbif` is developed
publicly on GitHub at <https://github.com/sckott/pygbif>, where
development versions of the package can be installed, and bugs and feature
requests reported. Stable versions of `pygbif` can be installed from
[pypi][pypipygbif], the distribution network for Python libraries.

@@

The output of these examples, except one, aren't shown.

@@

Spatial search, based on well known text format [@wkt], or a bounding box set of
four coordinates. The well known text string and the bounding box in the below
example specify the same rectangular area in California, centering
approximately on Sacramento. Whereas the bounding box format requires
`longitude SW corner, latitude SW corner, longitude NE corner, latitude NE corner`,
the well known text string requires an extra long/lat pair to close the polygon.

```{r eval=FALSE}
# well known text
wkt <- 'POLYGON((-122.6 39.9,-120.0 39.9,-120.0 37.9,-122.6 37.9,-122.6 39.9))'
occ_search(geometry = wkt, limit = 20)
# bounding box
occ_search(geometry = c(-122.6,37.9,-120.0,39.9), limit = 20)
```

Get only occurrences with lat/long data using the `hasCoordinate` parameter

@@

There's no equivalent interface in `pygbif` yet.

@@

# R vs. Python

Both R and Python are commonly used in science, and can be used for similar
tasks. Python, however, is a more general programming lanugae, and can be
used in more contexts than R can be used in.

The `rgbif` and `pygbif` clients can do a lot of the same tasks. We
envision `rgbif` as being more common in workflows of academics asking
research questions, whereas `pygbif` can do that as well, but may be more
easily used in a website.

The R client `rgbif` has much more development time than `pygbif`,
but through time `pygbif` will become equally mature.

@@

The `rgbif` and `pygbif` libraries provide programmatic interfaces to GBIF's
application programming interface (API) - a powerful tool for working with
species occurrence data, and facilitating reproducible research. In fact,
the `rgbif` package has already been used in 22 scholarly publications
(as of 2015-11-14).

@@

The `pygbif` library is in early development, and will
greatly benefit from any feedback and use cases.

One area of focus in the future is to attempt to solve many use
cases that have been brought up with respect to GBIF data. For
example, some specimens are included in GBIF that are located
in botanical gardens. For many research questions, researchers are
interested in "wild" type occurrences, not those in human curated
scenarios. Making removal of these occurrences easy would be
very useful, but is actually quite a hard problem. There are many
other problems like this.

@@

All scripts and data used in this paper can be found in the permanent
data archive Zenodo under the digital object identifier (DOI). This DOI
corresponds to a snapshot of the GitHub repository at
<https://github.com/sckott/msrgbif>. Software can be found at
<https://github.com/ropensci/rgbif> and <https://github.com/sckott/pygbif>,
both under MIT licenses. We thank all the users that have used `rgbif`
and `pygbif` and have given feedback and reported bugs. In addition, we
greatly appreciate the all contributions to the two packages, found
at <https://github.com/ropensci/rgbif/graphs/contributors> and
<https://github.com/sckott/pygbif/graphs/contributors>.
